Introducing Statistical Design of Experiments to SPARQL Endpoint Evaluation

### 2013-05-06 exp1/iswc2013.tex
Describe fractional experiments describe 32frac
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@516 43ef1190-90fd-4dd7-a769-d0121f02b80e

The main objective of this paper is not to establish that the
optimization is significant, it is merely a simulation. It is to
establish a critical practice of evaluations. It is of critical
importance that we understand why some effects are critical. That
``Implement'' is significant is hardly a surprise, that is what we
worked for. That the main effects ``Union'', ``Optional'' and ``Lang''
are strong is also due to that these are fairly demanding things to
evaluate. Since the delay we inserted affects all kinds of joins, it
is also intuitive that interactions between ``Implement'' and those
that require joins are strong. The strong ``BGPComp:Lang'' and
``Lang:Union'' interactions might be due to that the many more triples
needs to be searched for a language tag in one case. However, the
positive ``Implement:TripleC'' interaction evades such explanations;
in fact, it hints that our optimization may not work as well for
larger databases. This should cause some alarm. It is a fundamental
property of two-level experiments that it can only model linear
effects. If the effect is truly linear, and the variance is small, the
two levels can be quite close together and still yield a reasonable
estimate for the effect. However, if the effect is non-linear, the
choice of levels becomes critical. In this case, the experimenter
should be concerned that they may have a non-linear effect that is not
well represented by the sizes of datasets. Such reasoning should be
applied to all significant effects.

@@

>>>>

The main objective of this paper is not to establish that the
optimization is significant, it is merely a simulation. It is to
establish a critical practice of evaluations. It is of critical
importance that we understand why some effects are critical.\todo{add
  table of effects} That
``Implement'' is significant is hardly a surprise, that is what we
worked for. That the main effects ``Union'', ``Optional'' and ``Lang''
are strong is also due to that these are fairly demanding things to
evaluate. Since the delay we inserted affects all kinds of joins, it
is also intuitive that interactions between ``Implement'' and those
that require joins are strong. The strong ``BGPComp:Lang'' and
``Lang:Union'' interactions might be due to that the many more triples
needs to be searched for a language tag in one case. However, the
positive ``Implement:TripleC'' interaction evades such explanations;
in fact, it hints that our optimization may not work as well for
larger databases. This should cause some alarm. It is a fundamental
property of two-level experiments that it can only model linear
effects. If the effect is truly linear, and the variance is small, the
two levels can be quite close together and still yield a reasonable
estimate for the effect. However, if the effect is non-linear, the
choice of levels becomes critical. In this case, the experimenter
should be concerned that they may have a non-linear effect that is not
well represented by the sizes of datasets. Such reasoning should be
applied to all significant effects.

@@

As mentioned, the full factorial experiment goes as $2^n$ and becomes
prohibitivly expensive when $n$ is a large number of
factor. As the evaluation of SPARQL endpoints is an inherently complex
problem, a large number of factors are needed, and so full experiments
cannot scale. However, we may reduce the size of the experiment
significantly by sacrificing some explanatory power, in the form of
fractional factorial experiments.

We loose explanatory power due to \emph{aliasing}, for example, the
``TripleC:BGPComp'' interaction may be aliased with the
``Machine:Range'' interaction. That is to say, a detected increase in
run time from larger Basic Graph Patterns for large databases, can
also be explained by a less powerful machine evaluation FILTER clauses
with ranges. They are indistinguishable. This may or may not cause a
problem. In many cases, we may not be interested in these effects, we
may only be interested in ``Implement'' and its
interactions. Moreover, it is possible (even easy using the FrF2
package in R) to declare which effects must not be aliased, and
determine the size of the experiment based on that. Such a main effect
or two-factor interaction is called \emph{clear} if none of its
aliases are other main effects or two-factor interactions. Another
possibility to shrink the size of the experiment is if we \textit{a
  priori} can say that some effects are negligible. We have not found
such assumptions to be tenable in our case.

We have made two fractional factorial designs: One with 32 runs and
one with 64 runs. Again, the number of runs is in powers of 2, but
much smaller. In both cases, we have specified that all the factors
must be clear and also all two-factor interactions where ``Implement''
is involved must be clear. \todo{declare generators?}

The resulting normal plot is in \figure{fig:frac32normal}. We see that
only two main effects are significant on a $\alpha = 0.05$ level
according to the Lenth criterion, ``Implement'' is again deemed a
significant improvement, and ``Union'' is deemed significantly the
hardest operation for the query engine. We see that some other points
also deviate from the straight line, but with as few runs as this, the
total variance is great, so no other conclusions can be drawn. Also,
the Student's t-test we used in the previous section cannot be
employed, as we have no additional data to use for the test.

For any other purpose than a rough idea of some key influencers on the
overall performance of the endpoint, this little experiment is insufficient.
